Unifying Gompertzian growth with the communicable disease spreading paradigm

### 2022-04-25 main.tex
initial commit

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{float}
\usepackage{booktabs}       % professional-quality tables
\usepackage{graphicx}
\modulolinenumbers[5]

\journal{Chaos, Solitons \& Fractals}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
% \bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\insertPdfFig}[3]{
  \begin{figure}[H]
  \centering
  \includegraphics[width=12.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}

\newcommand{\insertSmallPdfFig}[3]{
  \begin{figure}[h]
  \centering
  \includegraphics[width=8.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}
\graphicspath{{../figures/}}
\begin{document}

\begin{frontmatter}

\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}

%% Group authors per affiliation:
% \author{Matz Andreas Haugen\fnref{myfootnote}}
% \address{Enerhauggata 1, 0651 Oslo, Norway}
% \fntext[myfootnote]{Since 1880.}

%% or include affiliations in footnotes:
\author[mymainaddress]{Matz Andreas Haugen\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
% \ead[url]{www.matzhaugen.com}
\ead{matzhaugen@gmail.com}

\author[mysecondaryaddress]{Dorothea Gilbert}

\address[mymainaddress]{Enerhauggata 1, 0651 Oslo, Norway}
\address[mysecondaryaddress]{University of Oslo, Oslo, Norway}

@@

\begin{keyword}
Gompertz \sep Coherence \sep Covid \sep Coronavirus \sep Network Analysis \sep Stochastic
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}

Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. Following this line of reasoning it can be shown that in the early stages of an epidemic, growth follows an initial exponential growth under a constant ratio of the infection rate and recovery rate. After the initial period, a logistic-like growth curve emerges, for which analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately,carletti2020covid}. However, instead of showing logistic-like growth, observed cumulative mortality exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
\begin{equation}
\label{eq:GompertzODE}
\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
\end{equation} 
% Redundant definitions for full script
with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
\begin{equation}
\label{eq:gomp_solution}
Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}}
\end{equation}
$Y_{0}=Y(t = 0)$, $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$ (see Fig. \ref{fig:Gompertz_concept} for a conceptual image). Rarely do macroscopic observations reliant on social networks behave so closely to a model \citep{broido2019scale}.

\insertSmallPdfFig{Gompertz_concept}{A conceptual display of the discrepancy seen between the logistic growth and the observations, superimposed by a Gompertz curve. Data is transformed to the straight line domain of the Gompertz curve for visual clarity, $f(Y)=\ln{\ln{1/Y}}$.}

@@

One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery ratio, $\phi(t)$. How would $\phi(t)$ have to behave? Start by recalling the SIR model for a pool of susceptible people of size $N$ evolving between the three states susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\\
\frac{dI}{dt}& = -\beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I,
\end{align}
omiting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

@@

 Due to this linearlization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions. And since $I=I(Y)$, we combine (\ref{eq:linSIR}) via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
\begin{subequations}
\label{eq:rypdal}
\begin{align}
\frac{dY}{dt} =& \gamma(t) Y(t) \label{eq:rypdalODE}\\
\gamma(t) =& \frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right] \label{eq:rypdalGamma},
\end{align}
\end{subequations}
where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$. This parametrization is the commonly used Richard's growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth. Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \citep{petroni2020logistic}. Note that at $\theta=1$, the traditional logistic growth curve is obtained, while at $\theta\rightarrow \infty$ we recover the exponential (Malthusian) explosion.

@@

 At this limit the growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. One would rather expect a near-constant relative growth rate in the early times. \citet{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries. Perhaps a more likely scenario is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. On the other hand, an infection of peripheral nodes should cause immediate exponential growth. Either case seems likely, and subject to initial conditions. \citet{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. We will touch upon how the Gompertz curve emerges from one such network below, under some caveats.

But even without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain growth which is almost Gompertzian, i.e. a straight line under a double-log transform of (\ref{eq:gomp_solution}), viz.
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow \citet{carletti2020covid} and consider the extented version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of ODEs (\ref{eq:SIR}), with a growth equation of the diseased group at some rate $d$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

Some algebraic manipulations reveal that the diseased group is described by a single ODE, viz.
\begin{equation}
\label{eq:SIRD_D}
\frac{dD}{dt} = \eta( 1 - e^{-\xi D}) - \kappa D,
\end{equation}
where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is assumed synonomous to the initial susceptible pool of people. This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $d$ and $a$. 
When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.

@@

With the full 3-parameter single ODE of the diseased group in the SIRD model, one can get almost arbitrarily close to the Gompertzian straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity. 
Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$, an effect which is manifested as a tapering of the initial straight line in the double-log domain.
These comparisons can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_ymax} and Supplemental Information).

Note also that the macroscopic SIRD model does not consider network effects but assumes rather all entities to be connected and communicating at each time-step, usually justified through mean-field approximations \citep{smilkov2014beyond}. 
It is however odd that observed mortality never exhibits a stage of near-exponential growth as this model predicts, but rather a constant slope in the double-log domain. We are thus prompted to look for another model which can explain the observations.

\insertPdfFig{Gompertz_vs_SIRD_lnln_ymax}{Cumulative diseased transformed by $f(Y)=\ln{\ln{({Y_{max} / Y(t))}}}$ plotted against number of days when $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text after normalizing, although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}). The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm. Each plot is annotated with an inferred transmission rate in the SIRD model, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

\section{An alternative theory for observed mortality}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability. This could explain the immediate dampening of mortality growth we see. 
Inspired by  \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term at the microscopic level along with the countering force of the immune system. 
As with all multiplicative processes, it is convenient to log-transform normalized mortality with variables $F(t)=Y(t)/Y_{\infty}$ and $Z=\ln{F}$, from which a natural perturbation model emerges,
\begin{equation}
\label{eq:microscopic}
dZ(t)= -\beta Z(t) dt + \sqrt{\sigma}dW(t),
\end{equation}
where $W(t)$ is a delta-correlated Wiener process\footnote{This perturbation model is also called an Ornstein–Uhlenbeck process}. Notice that this is just a continuous version of an AR(1) model. The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $K=\exp{(-\frac{\sigma}{2\beta})}$, viz. 
\begin{equation}
\label{eq:microscopicOriginal}
dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{K}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
\end{equation}

The first term on the right hand side represents the growth due to the stressor, while the second represents the stress response. 
Note that mortality is treated here as dimensionless where the microscopic treatment would interpret $F$ as cumulative probability of mortality, and the macroscopic equivalent would be the cumulative mortality.

To go from the microscopic to the macroscopic domain, we take the temporal average of (\ref{eq:microscopic}), and use the property that the average of log-quantities is the logarithm of the median quantity, to obtain the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}). We now see that by comparing (\ref{eq:rypdalODE}), (\ref{eq:rypdalLimit}) and the temporal average of (\ref{eq:microscopicOriginal}) that 
\begin{equation}
\sigma = 2\gamma_{\infty}\ln{(Y_{\infty})},
\end{equation}
showing the relationship between the diffusion coefficient and the final conditions.

Thus, a more direct interpretation of the observations is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate \citep{boxenbaum2017hypotheses,neafsey1988gompertz}. Here, a geometric rate is an exponential rate in the log-transformed domain, which is the natural transformation for many processes in nature \citep{zhang1994log}. The log-normal distribution of the abundance of $F$ can be obtained by solving the microscopic stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}. In fact, \citet{sitaram1984statistical} showed using methods from theormodynamics that a variable undergoing Gompertz growth is log-normally distributed in its abundance (see also \citet{gunasekaran1982lon}). This result is intuitively seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.

@@

\begin{align}
\label{eq:compare}
\begin{split}
\dot{F} = \frac{d}{dt}F(t) = \beta F (1-F) \quad&\quad \text{Logistic}\\ 
\frac{d}{dt}\ln{(F(t))} = -\beta \ln{(F(t)}) &\quad \text{Gompertz}
\end{split}
\end{align}

In the logistic model, we recognize the rightmost side of the logistic equation as the transmission term in an SIR model, but also as a linear interaction term between the two macroscopic states. Many studies do not consider microscopic effects at all, and the solution is obtained by solving the macroscopic (set of ODEs) \citep{vattay2020forecasting}.
This procedure is a mean field approximation with an implied average correlation between the variables. 
Thus, all dynamics are governed by macroscopic deterministic variables parametrized by a transmission rate.

Because the macroscopic logistic ODE solution never invokes the microscopic world, an imaginary microscopic solution could be modeled by splitting the system into $N$ units, $F(t) \rightarrow f_1(t), f_2(t),$ etc, from which macroscopic growth would be obtained by taking the average over the individual units,
\begin{equation}
F(t) = \frac{1}{N}\sum_i^N f_i(t).
\end{equation}

One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compare}), 
\begin{equation}
\label{eq:networkSIR}
\frac{d f_i}{dt} = \beta (1-f_i)\sum_j{a_{ij}}f_j \quad \forall i,
\end{equation}
with a fixed correlation governed by the network's growth rate, $b$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise. Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below. Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
\begin{equation}
\frac{\partial^2 \dot{F}}{\partial f_i \partial f_j} = -\frac{\beta}{N} a_{i,j}.
\end{equation}
Still, no Gompertz curve will emerge at the onset of the growth process. \citet{estrada2022networked} provide illuminating analysis on this topic.

>>>>

\documentclass{article}
% \documentclass[review]{elsarticle}

\usepackage{arxiv}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}   % Can be removed after putting your text content
\usepackage{graphicx}
\usepackage[square,numbers]{natbib}
\usepackage{doi}
\usepackage{amsmath}
\usepackage{float}

\newcommand{\insertPdfFig}[3]{
  \begin{figure}[H]
  \centering
  \includegraphics[width=8.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}

\newcommand{\insertSmallPdfFig}[3]{
  \begin{figure}[h]
  \centering
  \includegraphics[width=8.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}

% \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}
\graphicspath{{figures/}}
%\date{September 9, 1985} % Here you can change the date presented in the paper title
%\date{}          % Or removing it

\author{Matz A. Haugen, \\
Independent Reseacher, Oslo\\
\texttt{matzhaugen@gmail.com} \\
  \And
  Dorothea Gilbert \\
  University of Oslo\\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

% Uncomment to remove the date
%\date{}

% Uncomment to override  the `A preprint' in the header
%\renewcommand{\headeright}{Technical Report}
%\renewcommand{\undertitle}{Technical Report}
\renewcommand{\shorttitle}{On the nature of the Gompertz}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={A template for the arxiv style},
pdfsubject={q-bio.NC, q-bio.QM},
pdfauthor={Matz A. Haugen},
pdfkeywords={},
}

\begin{document}
\maketitle

@@

% keywords can be removed
\keywords{Gompertz \and Coherence \and Covid \and Coronavirus \and Network Analysis \and Stochastic}

\section{Introduction}
Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}. 
A large body of research has emerged that explores how these models fit the recent mortality seen due to the Coronavirus epidemic \citep{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.

However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
\begin{equation}
\label{eq:GompertzODE}
\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
\end{equation} 
% Redundant definitions for full script
with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
\begin{equation}
\label{eq:gomp_solution}
Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}}
\end{equation}
$Y_{0}=Y(t = 0)$, $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.

% \insertSmallPdfFig{Gompertz_concept}{A conceptual display of the discrepancy seen between the logistic growth and the observations, superimposed by a Gompertz curve. Data is transformed to the straight line domain of the Gompertz curve for visual clarity, $f(Y)=\ln{\ln{1/Y}}$.}

@@

One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$. How would $\phi(t)$ have to behave? Start by recalling the SIR model for a pool of susceptible people of size $N$ evolving between the three states susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\\
\frac{dI}{dt}& = -\beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

@@

 Due to this linearization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions. And since $I=I(Y)$, we combine (\ref{eq:linSIR}) via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
\begin{subequations}
\label{eq:rypdal}
\begin{align}
\frac{dY}{dt} =& \gamma(t) Y(t) \label{eq:rypdalODE}\\
\gamma(t) =& \frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right] \label{eq:rypdalGamma},
\end{align}
\end{subequations}
where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$. This parametrization is the commonly used Richard's growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \citep{wu2020generalized}. Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \citep{petroni2020logistic}. Note that at $\theta=1$, the traditional logistic growth curve is obtained, while at $\theta\rightarrow \infty$ we recover the exponential (Malthusian) explosion.

@@

 At this limit the growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
 The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. 
 One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
 \citet{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.

 Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
 On the other hand, an infection of peripheral nodes should cause immediate exponential growth. 
 \citet{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. 
 Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
 We will touch upon how the Gompertz curve emerges from one such network below, under some caveats.

Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain from the communicable disease models growth which is almost Gompertzian, i.e. a straight line under a double-log transform of (\ref{eq:gomp_solution}), viz.
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow \citet{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of equations (\ref{eq:SIR}), with a growth equation of the diseased group at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

Some algebraic manipulations reveal that the diseased group is described by a single equation, viz.
\begin{equation}
\label{eq:SIRD_D}
\frac{dD}{dt} = \eta( 1 - e^{-\xi D}) - \kappa D,
\end{equation}
where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is assumed synonymous to the initial susceptible pool of people. This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $\delta$ and $\alpha$. 
When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.

@@

With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity. 
Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters obtainable through linear regression, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$, an effect which is manifested as a tapering of the initial straight line in the double-log domain. 
Thus, through linear regression estimates, the Gompertz model avoids the possibility of non-identifiability issues and unstable optima \citep{roda2020difficult}.
Comparisons between the two models and the inevitable concavity of the SIRD model can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_ymax} and Supplemental Information).

Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond}. However, this model does not consider network effects explicitly. Rather, all entities are connected and communicating instantaneously as shown by \citet{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain. 
We are thus prompted to look for another model which can explain the observed mortality patterns.

\insertPdfFig{Gompertz_vs_SIRD_lnln_ymax}{Cumulative number of diseased, transformed by $g(Y)=\ln{\ln{({Y_{max} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD with a Gompertz model. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text after normalizing, although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}). The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm. Each plot is annotated with an inferred transmission rate in the SIRD model, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

\section{An alternative theory for observed mortality}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability. 
This could explain the immediate dampening of mortality growth observed. 
It could also explain the lack of correlation between population density and mortality (or infection) rates observed by some \citep{Hamidi2020,Hamidi2020a,Carozzi2020,Arpino2020,khavarian2021high,barak2021urban}.

Inspired by  \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term at the along with the countering force of the immune system. 
As with all multiplicative processes, it is convenient to log-transformed mortality with dimensionless variables $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
\begin{equation}
\label{eq:microscopic}
dZ(t)= -\beta Z(t) dt + \sqrt{\sigma}dW(t),
\end{equation}
where $dW(t)$ is a delta-correlated Wiener process with zero mean, making $Z(t)$ a stochastic process. This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}. It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}. The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $K=\exp{(-\frac{\sigma}{2\beta})}$, viz. 
\begin{equation}
\label{eq:microscopicOriginal}
dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{K}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
\end{equation}

The first term on the right hand side represents the growth due to the stressor, while the second represents the stress response.

To obtain the deterministic observable, we first take the average in the log-domain (\ref{eq:microscopic}) and transform back to the original domain, 
\begin{equation}
d\langle\ln{F(t)}\rangle = -\beta \langle\ln{F(t)}\rangle dt.
\end{equation}
Then use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $m(t)$, 
\begin{equation}
\label{eq:medianGomp}
d\ln{m(t)} = -\beta \ln{m(t)} dt,
\end{equation}
which corresponds with the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}) with $Y(t) = Y_\infty m(t)$. 
Comparing the stochastic stressor $\sigma$ in \ref{eq:microscopicOriginal} with the deterministic growth equation in (\ref{eq:GompertzODE}) gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor. 
We also see that by comparing (\ref{eq:rypdalODE}) and (\ref{eq:rypdalLimit}) with the stochastic counterpart in (\ref{eq:microscopicOriginal}) that the final growth level is governed by the stressor magnitude,
\begin{equation}
\sigma = 2\gamma_{\infty}\ln{(Y_{\infty})}.
\end{equation}

Thus, a more direct interpretation of the observations is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate \citep{boxenbaum2017hypotheses,neafsey1988gompertz}. 
Here, a geometric rate is an exponential rate in the log-transformed domain, which is the natural transformation for many processes in nature \citep{zhang1994log}. 
The log-normal distribution of the abundance of $F(t)$ can be obtained by directly solving the stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon}. 
Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.

@@

\begin{align}
\label{eq:compare}
\begin{split}
\dot{F} = \frac{d}{dt}F(t) = \beta F(t) (1-F(t)) \quad&\quad \text{Logistic}\\ 
\frac{d}{dt}\ln{(F(t))} = -\beta \ln{(F(t)}) &\quad \text{Gompertz},
\end{split}
\end{align}
where $F$ is here deterministic.

In the logistic model, we recognize the rightmost side of the logistic equation as the transmission term in an SIR model, but also as a linear interaction term between the two macroscopic states. 
As mentioned earlier, this procedure is a mean field approximation with an implied average interaction between the variables. 
Thus, all dynamics are governed by macroscopic deterministic variables parametrized by a transmission rate.

A microscopic solution could be modeled by splitting the system into $N$ microscopic deterministic units, $F(t) \rightarrow f_1(t), f_2(t),$ etc, from which macroscopic growth would be obtained by taking the average over the individual units,
\begin{equation}
F(t) = \frac{1}{N}\sum_i^N f_i(t),
\end{equation}
where the lower case $f_i$ emphasize the microscopic quality of the variables, and is here interpreted as probability of infection.

One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compare})
\begin{equation}
\label{eq:networkSIR}
\frac{d f_i}{dt} = \beta (1-f_i)\sum_j{a_{ij}}f_j \quad \forall i,
\end{equation}
using the shorthanded $f_i=f_i(t)$ and with a fixed correlation governed by the network's growth rate, $b$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise. Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below. Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
\begin{equation}
\frac{\partial^2 \dot{F}}{\partial f_i \partial f_j} = -\frac{\beta}{N} a_{i,j}.
\end{equation}
Still, no Gompertz curve will emerge at the onset of the growth process. \citet{estrada2022networked} provide illuminating analysis on this topic.

In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus suggests correlated mortality growth, viz.
\begin{equation}
F(t) = \exp{\left[\frac{1}{N}\sum_i^N z_i(t)\right]} = \left[\prod_i^N f_i(t)\right]^{1/N},
\end{equation}
where the rightmost equality is due to the central limit theorem where $z_i$ is a random variable as in \ref{eq:microscopic} while $f_i$ is here deterministic. 
Under this model, correlation between entities is present at all orders in the original domain\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - F(t))} = -b$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - F(t) = \exp{(-bt)}$ \citep{shklovskii2005simple}. 
In our context, the uncorrelated feature emerges since the force of mortality is not a function of the growth itself, as we see in the text, but rather of time.}.
